---
title: "CIND820_Capstone_EDA_Data_Cleaning"
output: html_document
date: "2025-06-17"
author: Elizabeth Esnard
---

Name: Elizabeth Esnard\
Course Section: CIND 820 TMU Student Number: 501342464

Load the raw data into a variable

```{r}
loan_working_copy <- read.csv("C:/Users/lizes/Documents/TMU - Chang School/CIND820/dataset bnpl/loan_working_copy.csv")
```

Check the summary to get a feel of the data.

```{r}
summary(loan_working_copy)
```

Confirm the number of rows in the raw data

```{r}
nrow(loan_working_copy)
```

Confirm the feature names

```{r}
colnames(loan_working_copy)
```

Filter the raw data into a subset that simulates BNPL purchase amounts.

```{r}
BNPL_clients <- loan_working_copy[(loan_working_copy$loan_amnt <= 3500), ]
nrow(BNPL_clients)
```

**Identifying the Target Variable.**

First, confirm all possible categories in the target variable feature.

```{r}
unique(BNPL_clients$loan_status)
```

We'll change all the categories into the two desired targets; solvent and default.

```{r}
library(dplyr)

BNPL_clients$loan_status <- recode(BNPL_clients$loan_status, "Current" = "solvent", "Fully Paid" = "solvent", "In Grace Period" = "solvent", "Late (16-30 days)" = "default")

BNPL_clients$loan_status <- recode(BNPL_clients$loan_status, "Late (31-120 days)" = "default", "Charged Off" = "default", "Default" = "default")
```

Confirm the distribution of the target variable. Both in a standard way (barplot function) and with a package to clean up the plot.

```{r}
barplot((table(BNPL_clients$loan_status)))
```

```{r}
library(ggplot2)

ggplot(BNPL_clients, aes(loan_status)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Distribution Of Target Variable",
       x = "Category",
       y = "Count") +
  theme_minimal()
```

There is a clear imbalance of the target variable, so there will be SMOTE oversampling to assist in the modeling.

Next, let's confirm the class type of all the features.

```{r}
sapply(BNPL_clients, class)

column_types <- data.frame(
  column = colnames(BNPL_clients),
  class = sapply(BNPL_clients, class)
)
```

Although there are a number of integer and/or numeric values, not all are truly numeric. Exploring the dataset in a spreadsheet, it's clear these values act as categories (e.g. 0 = no, 1 = yes) so through human logic I identified the true numeric features that summary statistics will reveal insightful information.

```{r}
BNPL_numeric <- BNPL_clients[c('loan_amnt', 'int_rate', 'monthly_payment', 'annual_inc', 'total_dti', 'tot_coll_amt', 'tot_cur_bal', 'credit_limit')]

summary(BNPL_numeric)
```

**Identifying Missing Values**

Now to identify any missing or null values that could affect models.

```{r}
colSums(is.na.data.frame(BNPL_clients))

```

```{r}
head(unique(BNPL_clients$emp_title), 20)
```

As seen through the unique function, as well as visual confirmation when browsing the data as a spreadsheet, the emp_title feature cannot easily be grouped or categorized. emp_title will not greatly contribute to model prediction, and will be removed from the working dataset.

```{r}
count(BNPL_clients[c((BNPL_clients$emp_title == ""),(BNPL_clients$emp_title == " ")), ])
```

R's is.na function is not identifying some null / missing values in emp_length and emp_title, which is why EDA was also performed in Python which came back with 7256 and 7812 respectively. Using the count function and sub-setting for specific empty cells ("") or single spaces (" ") returned the correct number of missing values in emp_length.

```{r}
nrow(BNPL_clients[BNPL_clients$loan_status == 'default',])

nrow(BNPL_clients[BNPL_clients$loan_status == 'solvent',])

nrow(BNPL_clients[is.na(BNPL_clients$total_dti) & BNPL_clients$loan_status == 'solvent', ])
```

Confirming is.na on total_dti, there are truly 36 observations with missing values in said feature.

```{r}
#remove 36 observations with total_dti missing
BNPL_clients <- BNPL_clients[!(is.na(BNPL_clients$total_dti) & BNPL_clients$loan_status == 'solvent'), ]

#remove emp_title column
BNPL_clients <- subset(BNPL_clients, select = -c(emp_title))

#replace NA values in emp_length
BNPL_clients$emp_length[BNPL_clients$emp_length == ' '] <- "5 years"
BNPL_clients$emp_length[(BNPL_clients$emp_length == 'n/a')] <- "5 years"
count(BNPL_clients[c((BNPL_clients$emp_title == ""),(BNPL_clients$emp_title == " ")), ])
```

Quickly we will check the distribution of emp_length with the filled in NA values.

```{r}
#barplot of emp_length distribution

#order the categories as factors so '10+ years" is at the end of the chart
BNPL_clients$emp_length <- factor(
  BNPL_clients$emp_length,
  levels = c("< 1 year", "1 year", "2 years", "3 years", "4 years", "5 years", "6 years", "7 years", "8 years", "9 years", "10+ years"),
  ordered = TRUE
)

ggplot(BNPL_clients, aes(emp_length)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Distribution Of Employment Length Variable",
       x = "Category",
       y = "Count") +
  theme_minimal()
```

```{r}
#barplot of loan_amount
ggplot(BNPL_clients, aes(x = loan_amnt)) + geom_histogram() +
  geom_density()
```

The data cleaning is now complete and can be exported.

```{r}
#export the file to working directory.

install.packages("openxlsx")
library(openxlsx)

write.xlsx(BNPL_clients, "clean_BNPL_clients.xlsx", rowNames = FALSE)
```

```{r}
write.csv(BNPL_clients, "clean_BNPL_clients.csv", row.names = FALSE)
```
